The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-16 EvolutionaryRescue1.tex
add fig

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{end float}
\usepackage{cleveref}
 \usepackage{subcaption}
\usepackage{xcolor}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

@@

% Daniel commands

\newcommand{\daniel}[1]{\textcolor{blue}{#1}}
\newcommand{\presc}{p_\text{rescue}}

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make more precise.
Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).

@@

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t>\left\lfloor\frac{3\log10}{\log\left(\frac{\lambda_m}{\mu_m}\right)}\right\rfloor+1.
\end{equation*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&\presc \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

\daniel{Need to explain intuitively what all this means.}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%\presc=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Density-dependent growth}

@@

In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
\presc 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

@@

\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{\presc(u>0)}{\presc(u=0)}.
\end{equation}

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

When aneuploid cells are partially resistant to the drug ($\Delta_w\ll$(SOMETHING $\ne0$)$\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}). 
When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll$(SOMETHING $\ne0$)$\ll\Delta_m$), it cannot rescue the population.
Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population. In this case, aneuploidy provides two benefits. First, it delays the extinction of the population--providing more time for appearance of the resistance mutations. Second, it increases the population size relative to a wildtype population--providing more cells for generating mutations, i.e., it increases the mutation supply.

@@

\section{Appendices}

\subsection*{Calculation of survival probabilities of single lineages}

Here we calculate the \emph{survival probability}, the probability that a lineage descended from a single cell does not become extinct.
Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a lineage started by a single wildtype cell, aneuploid cell, or mutant cell, respectively.
We will neglect density dependence, allowing us to treat the lineages as \emph{multitype branching processes} \citep{harris1963theory}.
The extinction probabilities $1-p_w$, $1-p_a$, and $1-p_m$ then satisfy the equations:
\begin{equation} \label{extinction_prob}
\begin{aligned}
1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
\end{aligned}
\end{equation}

\daniel{Rewrite the below in terms of $\Delta k$.}

The survival probabilities are given by the smallest solution for each quadratic equation \citep{weissman2009rate,uecker2015adaptive}. Therefore we have
\begin{equation}\label{survival_prob}
\begin{aligned}
p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
\end{aligned} 
\end{equation}
Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\Delta_k \equiv \lambda_k - \mu_k$ & 1/days & Growth rate of type $k$ cells, for $k=w,a,m$\\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

@@

%\presc=1-\e^{-\left(\theta_1+\theta_2\right)},

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{end float}
\usepackage{cleveref}
 \usepackage{subcaption}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

@@

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).

@@

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Survival probability}

To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
\begin{equation} \label{extinction_prob}
\begin{aligned}
1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
\end{aligned}
\end{equation}

The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
\begin{equation}\label{survival_prob}
\begin{aligned}
p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
\end{aligned} 
\end{equation}
Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.

@@

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Density-dependent growth.}

@@

In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

@@

\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
\end{equation}

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}). 
When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population. In this case, aneuploidy provides two benefits. First, it delays the extinction of the population--providing more time for appearance of the resistance mutations. Second, it increases the population size relative to a wildtype population--providing more cells for generating mutations, i.e., it increases the mutation supply.

@@

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
\caption{\textbf{TODO.}
TODO
}
\label{MeanTimeRatioInitialPopulationSize}
\end{figure}

@@

%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
